---
title: "Data Cleaning"
author: "Kim Donaldson"
date: "2/17/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
library(tidyverse)
```

## Part I

Data cleaning includes the following:

* Removing data we don't need for analysis
* Removing duplicate data
* Dealing with inconsistencies
* Dealing with missing data and outliers
* Creating new variables where necessary
* Combining seperate datasets

Questions to ask when cleaning data:

1. Which variables are needed for analysis?
2. Do we need to create any new variables from existing ones?
3. Are the columns the correct type?
4. Do we need to combine datasets?
5. Are there missing data? i.e NA, NULL, empty strings


```{r}
directory <- '/Users/kimdonaldson/Desktop/dataquest.io/data'
files <- c("sat_results", "ap_2010", "class_size", "demographics", "graduation_outcomes", "hs_directory")
filenames <- c()
for(file in files){
  file <- paste(directory, file, sep = "/")
  file <- paste(file, ".csv", sep = "")
  filenames <- c(filenames, file)
}
sat_results <- read_csv(filenames[1])
ap_2010 <- read_csv(filenames[2])
class_size <- read_csv(filenames[3])
demographics <- read_csv(filenames[4])
graduation <- read_csv(filenames[5])
hs_directory <-read_csv(filenames[6])
```

Useful functions for examining data include:

* `head()`
* `tail()`
* `str()`
* `colnames()`
* `glimpse()`
* `unique()`
* `map()`

```{r}
head(sat_results)
```

```{r}
colnames(sat_results)
```

We can get all the column data types by mapping the `class()` function to a data frame.

```{r}
map(sat_results, class)
```

Clean column names by following these steps:

1. Convert to lowercase
2. Remove punctuation
3. Convert spaces to underscores

```{r}
clean_col_names <- function(df){
colnames(df) %>% str_to_lower() %>%
  str_remove('[.]') %>%
  str_replace_all(" ", "_") %>% return()
}
colnames(sat_results) <- clean_col_names(sat_results)
colnames(ap_2010) <- clean_col_names(ap_2010)
colnames(class_size) <- clean_col_names(class_size)
colnames(demographics) <- clean_col_names(demographics)
colnames(graduation) <- clean_col_names(graduation)
colnames(hs_directory) <- clean_col_names(hs_directory)
colnames(sat_results)
```

We need to change the data types sat scores to numeric, and create a new variable that combines the scores

```{r}
sat_results <- sat_results %>%
  mutate(sat_critical_reading_avg_score = as.numeric(sat_critical_reading_avg_score),
         sat_math_avg_score = as.numeric(sat_math_avg_score),
         sat_writing_avg_score = as.numeric(sat_writing_avg_score),
         avg_sat_score = sat_critical_reading_avg_score + sat_math_avg_score + sat_writing_avg_score)
```

The `across()` function makes it easy to apply the same transformation. So we could have done the same type conversion as:
```{r}
sat_results <- sat_results %>%
  mutate( across( c('sat_critical_reading_avg_score','sat_math_avg_score','sat_writing_avg_score'), as.numeric ))
```

Fortunately we don't have to spell out the column names like this. `across()` also has helper functions `starts_with()`, `ends_with()`, and `contains()`
So we could have just as easily written this out as:
```{r}
sat_results <- sat_results %>%
  mutate( across(contains('avg_score'), as.numeric) )
```

```{r}
colnames(ap_2010)
```

```{r}
map(ap_2010, class)
```

```{r}
ap_2010 <- ap_2010 %>%
  mutate(
    exams_per_student = total_exams_taken / ap_test_takers,
    high_score_percent = (number_of_exams_with_scores_3_4_or_5 / total_exams_taken) * 100
  )
```

Now lets look at class_size dataframe
```{r}
colnames(class_size)
```

```{r}
print(unique(class_size$grade))
print(unique(class_size$program_type))
```
Instructions:

1. Filter class_size where grade values are equal to '09-12' and program type values are 'GEN ED'

```{r}
class_size <- class_size %>%
  filter(grade == '09-12' & program_type == 'GEN ED')
```

Create a new class_size DataFrame that contains the following:

The following variables from the original DataFrame:

CSD
SCHOOL CODE
SCHOOL NAME
The following new variables:

avg_class_size (average of AVERAGE CLASS SIZE for each school)
avg_largest_class (average of SIZE OF LARGEST CLASS for each school)
avg_smallest_class (average of SIZE OF SMALLEST CLASS for each school)

```{r}
class_size <- class_size %>%
  group_by(csd, school_code, school_name) %>%
  summarise(avg_class_size = mean(average_class_size),
            avg_largest_class_size = mean(size_of_largest_class),
            avg_smallest_class_size = mean(size_of_smallest_class))
```

The graduation data frame:
Create a new graduation DataFrame.

Filter the rows that contain only the following:

Data for the total cohort of students (the Demographic variable is equal "Total Cohort") and
Data from the 2006 Cohort (the Cohort variable is equal "2006").
Select the following columns : DBN, School Name, Total Grads - % of cohort, Dropped Out - % of cohort.

```{r}
graduation <- graduation %>%
  filter(demographic == 'Total Cohort' & cohort == '2006') %>%
  select(dbn, school_name, `total_grads_-_%_of_cohort`, `dropped_out_-_%_of_cohort`)
```

Create a new demographics DataFrame.

Filter the rows that contain only the following:

Data for which values of schoolyear are "20112012" and
Data for which values of grade9 are not "NA" (i.e., ! is.na(grade9)).
Select Columns containing the following variables using the fact that some of them contain "_per" in their name:

DBN
Name
frl_percent
total_enrollment
ell_percent
sped_percent
asian_per
black_per
hispanic_per
white_per
male_per
female_per

```{r}
demographics <- demographics %>%
  filter(schoolyear == '20112012' & !is.na(grade9)) %>%
  select(dbn, name, total_enrollment, contains('_per'))
```

Create a new demographics_clean DataFrame by removing the Name and the female_per columns from the demographics DataFrame available from the previous screen.

Use the negative column technique to remove the Name and the female_per columns.
Save the new DataFrame as demographics_clean.

```{r}
demographics_clean <- demographics %>%
  select(-name, -female_per)
```

Create a new hs_directory DataFrame that contains columns for the variables dbn, school_name, and Location 1.

```{r}
hs_directory <- hs_directory %>%
  select(dbn, school_name, location_1)
```

## Part 2 String manipulation

Tidy Data:

* Each variable must have it's own column name
* Each observation mush have it's own row
* Each value must have it's own cell

The `parse_number()` function takes a vector as input and drops and characters before and after the numeric part of a string. This can be used in combination with `mutate()` and `across()` 

df %>% 
  mutate( across(first_col:last_col), parse_number )

Let's look at the `hs_directory` dataset. The location_1 column contains two '\n' This information can be separated into it's own columns with the `separate()` function with the syntax:

df %>% separate(col = name_of_col_to_separate, 
                into = c("col1", "col2", "col3"),
                sep = '\n')
                
Instructions:
Create a new hs_directory dataframe containing a new variable column, lat_long, containing the coordinates (surrounded by parentheses) extracted from the Location 1 column.

Use the separate() function to split the column Location 1 into three new columns: "string_1", "string_2", and "string_3" with "\n" as separator.
Use the select() function to remove the columns string_1 and string_2 because we don't need them.
Use the rename() function to rename the string_3 column into lat_long.

```{r}
hs_directory <- hs_directory %>%
  separate(col = location_1,
           into = c("string_1", "string_2", "string_3"),
           sep = '\n')
```
```{r}
hs_directory <- hs_directory %>% select(-string_1, -string_2)
```
```{r}
hs_directory <- rename(hs_directory, lat_long = string_3)
```

Instructions:

Create a new hs_directory dataframe containing new columns, lat and long, containing the coordinates extracted from the lat_long column.

Use the separate() function to split the column lat_long into two new columns: "lat" and "long" with "," as the separator.

```{r}
hs_directory <- hs_directory %>%
  separate(col = lat_long,
           into = c("lat", "long"),
           sep=',')
```

Instruction:
Use str_sub() and mutate() to create a new hs_directory dataframe with lat and long variables containing only numeric characters.

Convert the data type of the lat and long variables from character to numeric.

For this exercise, we can also use parse_number() to achieve the same result. Don't hesitate to try it!

```{r}
hs_directory <- hs_directory %>%
  mutate(lat = parse_number(lat),
         long = parse_number(long))
```

`str_c()` concatenates strings
`str_pad()`, adds padding

Create a new DBN variable in the class_size dataframe.

Combine the columns CSD and SCHOOL CODE and assign the result to a new column DBN.
Update the DBN column making sure that its values have six characters.

```{r}
class_size <- class_size %>%
  mutate(dbn = str_c(csd, school_code, sep = "")) %>%
  mutate(dbn = str_pad(dbn, width = 6, side = 'left', pad = '0'))
```

## Part 3 - Relational Data

The `duplicated()` function takes a vector as input and returns a logical vector that indicates wether the value is a duplicate of the one that comes before it. This can be combined with `sum()` to get a count of duplicated rows. In order to remove duplicated rows we can use the `dplyr` function `distint()` with the syntax:
df %>% distinct(col_name, .keep_all = TRUE)

There are two main types of mutating joins

* inner joins
  Matches pairs of variables in two dataframes whenever their values of keys are the same. Does not include any rows with unmatched keys
* outer joins
  Keep rows that appear in atleast one of the two dataframes we're combining. Can be divided into left, right, and full joins

Instructions:
Use the `inner_join()` function to combine the sat_results and class_size dataframes.

Save the new dataframe as sat_class_size.

Create a scatter plot with avg_class_size on the x-axis and avg_sat_score on the y-axis to see if SAT scores seem to be related to class size.

```{r}
sat_class_size <- sat_results %>%
  inner_join(class_size, by = 'dbn')

ggplot(data = sat_class_size, aes(x = avg_class_size, y = avg_sat_score)) + 
  geom_point()
```


```{r}
combined <- sat_results %>%
  full_join(ap_2010, by = "dbn") %>%
  left_join(class_size, by = "dbn") %>%
  left_join(demographics, by = "dbn") %>%
  left_join(graduation, by = "dbn") %>%
  left_join(hs_directory, by = "dbn")
```

