---
title: "Learning Plotly"
author: "Kim Donaldson"
date: "1/1/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


## Introduction to graphing with Plotly

```{r}
library(tidyverse)
library(plotly)
library(palmerpenguins)
vaers_file <- '/Users/kimdonaldson/Desktop/2021VAERSData/2021VAERSDATA.csv'
vaers_data <- read.csv(vaers_file)
vaers_deaths <- vaers_data %>%
  filter( DIED == 'Y' )
vaers_deaths_25_50 <- vaers_deaths %>%
  filter( AGE_YRS < 50 & AGE_YRS > 25)
```
## Basic Scatter Plot
Comparing Body Mass vs. Flipper length in Penguins

```{r}
fig <- penguins %>%
  plot_ly( x = ~body_mass_g, y = ~flipper_length_mm, type="scatter", color = ~species) %>%
  layout(
    title = "Body Mass vs. Flipper Length",
    xaxis = list(title="Body Mass"),
    yaxis = list(title="Flipper Length")
  )
fig
```
## Basic Histogram in ggplot
```{r}
ggplot(data=penguins, aes(x = body_mass_g)) + geom_histogram(
  
)
```


Basic Histogram in Plotly
```{r}
fig <- penguins %>%
  plot_ly( x = ~body_mass_g, type="histogram", color = ~body_mass_g, colors=c(rgb(.8,0,0)), opacity=.5) %>% 
  layout(
    title = "Body Mass"
  )
fig
```

```{r}
fig <- vaers_deaths %>% 
  plot_ly(x = ~CAGE_YR, type="histogram", color = ~AGE_YRS, colors="dark blue", legend = "none", name="Vaers Deaths") %>%
  layout(
    title = "VAERS Deaths in Years",
    xaxis = list(title = "Age"),
    yaxis = list(title = "Count"),
    plot_bgcolor="#c7daec",
    showlegend = FALSE
  )
fig
```

```{r}
x <- seq_len(nrow(volcano)) + 100
y <- seq_len(ncol(volcano)) + 500
plot_ly() %>% add_surface( x = ~x, y = ~y, z=~volcano)
```
## Bar Charts
```{r}
fig <- plot_ly(diamonds, x = ~cut, color=~clarity, colors = "Greens", stroke = I("black"))
fig
```

Overlaying Histograms
```{r}
a <- penguins %>% filter(species=="Gentoo") %>% select(body_mass_g)
b <- penguins %>% filter(species=="Adelie") %>% select(body_mass_g)
c <- penguins %>% filter(species=="Chinstrap") %>% select(body_mass_g)
fig <- plot_ly(opacity=.6, stroke = I("black")) %>%
  add_histogram(data = a, x=~body_mass_g, name="Gentoo") %>%
  add_histogram(data = b, x=~body_mass_g, name="Adelie") %>%
  add_histogram(data = c, x=~body_mass_g, name="Chinstrap")
fig <- fig %>% layout(barmode="overlay", title="Penguin Body Mass by Species",
                      xaxis=list(title="Body Mass"),
                      yaxis=list(title="Frequency"),
                      plot_bgcolor = "#c7daec",
                      showlegend = TRUE)
fig
```

Creating a Bar Chart of the average body mass of penguin species
```{r}
penguin_stats <- penguins %>% 
  group_by(species) %>% 
  drop_na() %>%
  summarise(
    min_body_mass = min(body_mass_g),
    avg_body_mass = mean(body_mass_g),
    max_body_mass = max(body_mass_g))
fig <- penguin_stats %>% plot_ly( x = ~species, y = ~avg_body_mass, color = ~species) %>%
  add_bars() %>%
  layout(title="Average Body Mass",
         xaxis=list(title="Species"),
         yaxis=list(title="Body Mass in Grams"),
         barmode="stack")
fig
```

We can see that Adelie and Chinstraps are about the same size, and Gentoo's are about 27% bigger. 
```{r}
fat_penguin <- function(weight, threshold = 4177){
  if(weight > threshold){
    print("Yeah...maybe")
  } else {
    print("Maybe not")
  }
}
```

Basic bar plot

```{r}
nums <- c(100, 200, 300)
names <- c("A", "B", "C")
bar_df <- data.frame(nums, names)
```


