---
title: "Linear Algebra"
author: "Kim Donaldson"
date: "1/6/2022"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Solving Linear Systems

1. Get a variable by itself in one of the equations
2. Substitue the variable from step 1 into the other equation
3. Solve the equation in step 2 for the remaining variable. 

## Row Echelon Form

1. All pivot entries are equal to 1
2. Any rows that consist of only 0's are at the bottom
3. The pivots form a stair case pattern descending to the right

## Reduced Row Echelon Form

4. Every pivot is the only non zero entry in it's vector

## Guass Jordan Elimination

1. Pull out scalars from each row in the matrix
2. If the first entry in the first row is 0, swap it for with another row that has a non zero entry in it's first position
3. Multiply through the first row by a scalar to make the leading entry equal to 1
4. Add scaled multiples of the first row to every other row until every entry in the first column, other than the leading 1 in the first row, is a 0
5. Go back to Step 2 and repeat

Implemntation 
```{r}
find_gcd <- function( x, y){
  r <- x %% y
  return(ifelse(r, find_gcd( y, r), y))
}

gcd <- function(num_vector = c()){
  if (mode(num_vector) != "numeric"){
    return("gcd error: vector must be numeric")
  }
  else if(min(num_vector) < 0){
    return("gcd error: all values must be positive")
  }
  else {
    num_vector <- sort(num_vector)
    r <- NULL
    prev_num <- NULL
    c <- 0
    
    vec_len <- length(num_vector)
    for(num in num_vector){
      c <- c + 1
      if(c == 1){
        next
      } else {
        prev_num <- num_vector[c-1]
        r <- append(r, find_gcd(prev_num, num))
        #print(r)
      }
    }
  }
return(find_gcd(min(r), max(r)))
}
```
```{r}

k.make <- function(){
  return( data.matrix( data.frame( c(1:6)*2, c(1:6)*3, c(1:6)*5 )))
}

# Overwrites default range() 
range <- function(n){
  return(c(1:n))
}

k.reduce <- function(X){
  for(i in range(ncol(X))){
    s <- gcd( X[,i])
    if(s > 1){
      X[,i] <- X[,i] / s
    }
  }
  negative_row <- X[,1] * -1
  
  # Check if the first pivot entry is greater than 1, if so normalize the first vecor such that the first entry 
  # becomes a 1
  if(X[X[1,1]] > 1){
    X[,1] <- X[,1] / X[X[1,1]]
  } else {
    # Otherwise it's already a 1 and we can move on
    print("Debug")
    print(X)
    for(i in range(ncol(X))){
      if(i == 1){ next }
      else {
        X[,i] <- X[,i] + negative_row
      }
    }
  }
  return(X)
}

k.display <- function(X){
  for(i in range(ncol(X))){
    print(X[,i])
  }
}


```

```{r}
for(i in range(ncol(K))){
  s <- gcd(K[,i])
  K[,i] <- K[,i] / s
}
```


