---
title: "R4DS"
author: "Kim Donaldson"
date: "1/9/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(palmerpenguins)
library(plotly)
vaers_file <- '/Users/kimdonaldson/Desktop/2021VAERSData/2021VAERSDATA.csv'
vaers_data <- read.csv(vaers_file)
vaers_deaths <- vaers_data %>%
  filter( DIED == 'Y' )
```


## Data Visualization with ggolot and plotly

### Creating a Scatter Plot to show how two variables are correlated.
Examine the trend between cars engine size on highway mpg.
In general, we put the explanatory variable on the X axis, 
and the response variable on the Y axix.

```{r}
ggplot( data = mpg, mapping = aes(x = displ, y = hwy )) +
  geom_point() +
  geom_smooth()
```

This shows that the larger the engine size, the smaller the fuel effeciency. 
We can also wrap ggplot objects in plotly to make charts look nicer. 
```{r}
g <- ggplot(data = mpg, mapping = aes( x = displ, y = hwy, color=class)) +
  geom_point() 

ggplotly(g) %>%
  layout(
    title="Engine Size vs. Fuel Efficiency",
    xaxis = list(title="Engine Size"),
    yaxis = list(title="Miles Per Gallon"),
    plot_bgcolor = 'b0fafc'
  )
```
Excercises
1. What's gone wrong with this code? Why are the points not blue?
```{r}
ggplot(data = mpg) +
  geom_point( 
    mapping = aes( x = displ, y = hwy, color="blue" )
    )
```
Solution: the color attribute should be inside the geom_point() function,
not inside the mapping attribute:
```{r}
ggplot(data = mpg) +
  geom_point( 
    mapping = aes(x = displ, y = hwy), color="blue")
```
2. Which variables in mpg are categorical? Which are continuous?
Categorical: manufacturer, model, trans, drv, fl, class
Continuous: displ, cty, hwy

3. What does the stroke aesthetic do? 
Stroke adjusts the thickness of the border for shapes that can take an
interior and exterior color. It only works on shapes 21-24

## Coordinate Systems

* coord_flip() switches the x and y axes
```{r}
ggplot(data = mpg, mapping=aes(x = class, y = hwy)) +
  geom_boxplot() +
  coord_flip()
```

## Statistical Transformations

```{r}
ggplot(data = diamonds) +
  geom_bar( mapping = aes(x = cut, fill=cut) )
```

### Layered Grammar of Graphics

#### GGPLOT DATA = DATA
##### GEOM_FUNCTION
##### Mapping = aes mappings
##### stat = statistical transformation
#### COORD FUNCTION
#### FACET FUNCTION

## Chapter 2: Data Transformation

```{r}
library(nycflights13)
```

### dplyr basics

* filter() select obs by their values
* arrange() reorder the rows by ascending or descending order
* select() pick variables by name
* mutate() create new variables based on the values of other variables
* summarize() create a table of summary statistics
* group_by() change the scope of each function to operate on the dataset by group

