---
title: "Learning Journal"
subtitle: "Statistics and Probabilities with R"
last-update: "6/12/2022"
author: "Kim Donaldson"
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

## Statistics with R

**Data Basics**

* Types of variables
* Relationship between variables
* Questions about variables
1. What type of variation occurs between variables
2. What type of covariation occurs

### Variables

* **Numerical**
  + **Continuos** 
    - Infinite values within a given range
  + **Discrete**
    - Take on one of a specifit set of numerical values
* **Categorical**
   + **Regular Categorical**
   + **Ordinal**
     - Levels have an inherent ordering
     
## Vizualizing Numerical Data

Generally plot the explanatory variable on the x axis and the response variable on the y axis.
```{r}
library(tidyverse)
library(palmerpenguins)
library(plotly)
```

Graph of a scatter plot showing the explanatory variable on the X axis and the response variable on the Y axis

```{r}
ggplot(data = penguins, aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_point() +
  geom_smooth()
```
```{r}
ggplot(data = penguins, aes(x = body_mass_g)) +
  geom_histogram()
```

### Estimates of distribution

* mean()
  + weighted.mean()
  + trimmed mean mean(list, trim=.1)
* median()
* sd()
* mad()
* range()
* IQR()


## Normal Distribution

Also referred to as a bell curve. 
Unimodal and symtetric

### 68 - 95 - 99.7% Rule

When a numerical variable is normally distributed 68% of the values fall within 
one standard deviation of the mean. 95% falls within 2 standard deviations, and 
99.7% falls within 3 standard deviations

The __rnorm()__ function will generate random numbers that fall within a normal distribution.

```{r}
nx <- rnorm(200)
ndata <- data.frame(x = nx)
ndata %>% plot_ly( x = ~x, type = 'histogram', color = ~x, colors ='dark green') 
```

### Standardized Z Scores

Standardized (Z) score of an observation is the number of standard deviations
it falls above or below the mean. 

Z = ( observation - mean ) / standard deviation

A Z score of mean = 0

A Score of |2| is considered to be an unusual observation


## Binomial distribution

Describes the probability of havinv exactly k successes in n independent trials
with probability of success p.

__Bernoulli Random Variable__: An individual trial with only two possible outcomes

__Binomial Distribution__ the probability of having exactly K successes in N 
independent trials with probability of success p. 

If p represents the probability of success, (1-p) represents the probability of 
failure, n represents the number of independent trials, and k represents the number of 
successes. 

P(k successes in n trials) = (n choose k) * p^k * (1 - p)^(n - k) 

__Binomial Conditions:__

1. Trials must be independent
2. The number of trials (n) must be fixed
3. Each outcome must be classified success or failure
4. The probaibily of success (p) must be the same for each trial 

__# of scenarios * P(single scenario)__

The __choose()__ function can be used to calculate the number of ways to choose K 
successes in N trials. 

(N choose K) is calculated as: N! / K! * (N - K)!
This can be done in R with the dbinom() function

How many scenarios yield three successes in five trials? 
N = 5, K = 3
```{r}
choose( 5, 3 )
```
How many scenarios yield 2 successes in nine trials?
N = 9, K = 2
```{r}
choose(9, 2)
```

If P represents success, then (1 - P) represents the probability of failure,
N reprsents the number of trials, and K reprsents the number of successes.

P(K successes in N trials) = N choose K *  P^k * (1-P)^(N-K)

This can be calculated with the __dbinom(K, N, P)__ function

According to a 2013 Gallop poll, worldide only 13% of empployees are engaged at work.
Among a random sample of 10 employees, what is the probability that 8 of them are engaged at work?

```{r}
dbinom(8, size = 10, p = 0.13 )
```

The __pnorm()__ function calculates the percentile of where a number falls in
the normal distribution given the mean and standard deviation of the distribution.

Example: SAT scores distributed normally with mean 1500, and SD 300.
If a test taker scores 1800 we can find the percentile with:
```{r}
pnorm(1800, 1500, 300)
```

The __qnorm(percentile, mean, sd)__ function caclulates the minimum number of a percentile
given the percentile, mean, and standard deviation. 

Example: An SAT test taker tells you that they scored in the top 10%.
What is the minimum score they could have received?
```{r}
qnorm( 0.90, 1500, 300 )
```


Plotting a binomial distribution.

```{r}
num_trials <- 50
p <- .5
y <- c()
x <- c()

for(i in c(1:num_trials)){
    p_val <- dbinom( i, num_trials, p = p ) 
    y[i] <- p_val
    x[i] <- i 
}


df <- data.frame(x,y)
df %>% plot_ly(x = ~x, y = ~y, type='bar', color=~y, colors="Oranges") %>%
  layout(
    title="Normal Distribution"
  )
```

## Set Theory

A __Set__ is a collection of elements. 
The __Cardinality__ of a set is the number of elements in a set.
The __Intersect__ of a number of sets is the elements that the sets have in common.
The __Union__ of a number of sets is the elements that are in either set. 

## Probability

The probability of an event is a number between 0 and 1 where 0 represents that 
an outcome is not true, and where 1 represents that an outcome is true. 

__Independence__: Two events are independent if knowing the outcome of one event does not change the outcome
of the probability of the other. 
Notation P(A|B) = P(A)

__Joint Probability__ is the probability that two seperate independent events are both true.
Notation: P( A,B ) = P(B) * P(A)
Example: Two coins are tossed at the same time. What is the probability that both will land heads?
P(H) = 0.5
P( H,H ) = 0.5 * 0.5 = 0.25

__Disjoint Events__: Events that are mutually exclusive.
Notation: P( A or B ) = P(A) + P(B)

__General Addition Rule__: P( A or B ) = P(A) + P(B) - P(A and B)
Example: What is the probability that a randomly chosen card is a face card or a Heart?
These events are not disjoint since they can both occur, so we have to subtract the union.
P(Hearts or face card) = P(Hearts) + P(face card) - P(Hearts and face card)
P(Hearts or face card) = 13/52 + 12/52 - 3/52

__Permutation__: A set of elements that follows a linear order.
Formula: N! / (N - K)! Where N is the number of unique objects and K is the number of unique attributes
Example: There are 12 runners in a race. There are three unique attributes 1st place, 2nd place, 3rd place.
How many permutations are possible?
```{r}
factorial(12) / factorial(12 - 3) 
```

__Combination__: A set of elements where order does not matter.
Formula: N! / K! * (N -K)!
This is the same as N choose K

__Marginal Probability__: The probability that an individual event from one experiment occurs, regardless of the outcome from another experiment. 
Notation: P(A1)

__Conditional Probability__: When one event impacts the probability of another event.
Notation: P(A|B) probability of A given event B

__Product Rule__: P(A|B) = P(A, B) / P(B)

## Bayes Rule

P(A|B) = P(B|A) * P(A) / P(B)
Where A is a hypothesis and B is observed data
